A family of accessibility measures (bringing balance back?)

### 2025-01-03 paper/manuscript.qmd
Cleaned up repo - latest version of the manuscript draft.

---
title: "A family of accessibility measures"
format: pdf
bibliography: "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
---

@@

```{r load-packages, include=FALSE, cache=FALSE}
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(flextable)
# library(huxtable) # Easily Create and Style Tables for LaTeX, HTML and Other Formats
# library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax
# library(patchwork) # The Composer of Plots
# library(RColorBrewer) # ColorBrewer Palettes
# library(sf) # Simple Features for R
# library(scales) # Scale Functions for Visualization
# library(shadowtext) # Shadow Text Grob and Layer
# library(skimr) # Compact and Flexible Summaries of Data
# library(spdep) # Spatial Dependence: Weighting Schemes, Statistics
# library(tidyr) # Tidy Messy Data
# library(tmap) # Thematic Maps
# library(ggpmisc) # Miscellaneous Extensions to 'ggplot2'
# library(ggrepel) # Automatically Position Non-Overlapping Text Labels with 'ggplot2'
# library(ggspatial) # Spatial Data Framework for ggplot2
# library(gridExtra) # Miscellaneous Functions for "Grid" Graphics 
# library(here)
```

@@

\noindent where $B_a$ is the amount of business drawn to $a$ from $T$, $P_a$ and $P_T$ are the populations of the two settlements, and $D_{aT}$ is the distance between them. Quantity $N$ was chosen by Reilly in a somewhat _ad hoc_ fashion as 1, and he used empirical observations of shoppers to choose a value of $n = 2$.

@@

The entropy maximization derivation is demonstrated in detail elsewhere (e.g., @ortuzar_2011_modelling, @wilsonSTATISTICALTHEORYSPATIAL1967). The basis of the approach is to accept that there are macro, meso, and micro states. In the case of trips being the interaction variable, a micro state is an individual trip maker, the meso state is the statistically representative zone representing a collection of individual trip makers, and the macro state is the entire study region. Using these assumptions, a set of $T_{ij}$ that maximizes the number of micro states in each meso state subject to some sensible macro-state constraints (@eq-constraint0_5-gravitymodel, @eq-constraint1-gravitymodel, and @eq-constraint2-gravitymodel). From this approach, the gravity model can be interpreted as the given the total number of individual trips (micro level) statistically represented at origin and destination zones (meso level), the cost of travelling (micro level, represented at the meso level), and the fixed total cost of travelling in the region (number of trips and cost of travel at the macro level). There is then a most probable distribution of trips between zones (meso-states) that stasis all the macro-state constraints.

@@

And finally, Wilson's balancing factors are interpreted as inverse Hansen-type accessibility, especially in the context of competition [@vickermanAccessibilityAttractionPotential1974a; @karstEvaluationAccessibilityImpacts2003;@geurs2006accessibility; @willigers2007accessibility;@el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling]. This interpretation is fraut, upon considering @fotheringhamSpatialCompetitionAgglomeration1985's demonstration, which many of these works do not contend with.

@@

$$
V_{ij} = k W_j^{(2)} f(c_{ij})
$${#eq-access}

Where $V_{ij}$ consists of the same variables, however $V_{ij}$ explicitly measures _potential interaction_ as a unit of number of potentially reachable opportunities and the single proportionality constant $k$ ensures the total sum of $V_{ij}$ represents the total potential reachable opportunities in the data. Accessibility is also most often presented as a locational measure, the value of $V_i$ is of most interest, e.g., the number of potentially reachable opportunities at a zone.

Like with @eq-phys-gravity-model, we distinguish @eq-access into four cases based on versions of the gravity-model building constraints (@eq-constraint1-gravitymodel,  @eq-constraint2-gravitymodel, and @eq-constraint3-gravitymodel) to reflect macro-level knowledge, i.e., the sum of the number of potential reachable opportunities.

@@

All four <!--five?--> cases are differentiated like:
- The unconstrained case: neither @eq-constraint0_5-gravitymodel, @eq-constraint1-gravitymodel, or @eq-constraint2-gravitymodel hold;
- The total constrained case: only @eq-constraint0-gravitymodel hold
- The singly-constrained case: either @eq-constraint1-gravitymodel holds (origin-constrained) <!-- production--> or @eq-constraint2-gravitymodel holds (destination-constrained)<!-- attraction-->; 
- The doubly-constrained case: both @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel hold simultaneously <!--production-attraction-->
- HYBRID CASE: @eq-constraint1-access or @eq-constraint2-access partly hold -- hybrid production constrained case <!-- (could be repeated for attraction constrained case... and doubly constrained case)-->

We can use the same trip matrix, but reformulate its meaning. Let's say $W_i^{(1)}$ is population at $i$ who seek opportunities, lets recast this variable as $P_i$. Then let's say $W_j^{(2)}$ are the opportunities being sought, recast to $O_j$. For this simple example, we can say that the opportunities are the population's jobs. In our region, each person has a job -- but we only know at an aggregate how many people are at each $i$, the jobs that are located at $j$ and the travel cost ($C$ and associated travel propensity $1/C$). (let's pretend our 'known_trip_matrix' is unknown).

```{r}
known_trip_matrix <- matrix(c(95,23,42, 27,378,45, 38,36,106), ncol = 3, byrow = TRUE)
#unknowns <- matrix(c(rep(NA,9)), ncol=3, byrow=TRUE)
W_i <- known_trip_matrix |> rowSums()
W_j <- known_trip_matrix |> colSums()
T <- ifelse(sum(W_i)==sum(W_j), sum(W_j),NA)
knowns_longer <- cbind(known_trip_matrix,W_i)
knowns_longer <- rbind(knowns_longer,cbind(t(W_j),T))

@@

- No proportionality constant, $k$ is set to 1, so no constraints hold, making the interpretation of the units meaningless so the value's magnitude can only be interpreted:
$$
V_{ij} = k O_j f(c_{ij})
$$

- In practice, $V_i$ is also calculated for each $i$. When thinking about the interpretation of the value, how many opportunities can be visited for each $i$, the $P_i$ is also set to 1:
$$
V_{i} = \sum_j O_j f(c_{ij})
$$
These numbers are proportional to the accessibility (i.e., higher or lower), but are not scaled to number opportunities (jobs).

Looks like Hansen, but multiplied by P_i.

@@

This needs a balancing factor, $k$ is replaced with $K$:
$$
V_{ij}' = K P_i O_j f(c_{ij})\\
V_{i}' = K P_i \sum_j O_j f(c_{ij})
$$

This approach is not seen in accessibility literature...

@@

Similarly, singly-constrained can be either origin-constrained $V_i''^{(1)}$ or destination-constrained $V_i''^{(2)}$. For the origin- and destination- constrained versions, $k$ is replaced with $A_i$ and $B_j$ respectively:
$$
V_{ij}''^{(1)} = A_i O_j f(c_{ij})\\
V_{ij}''^{(2)} = B_j O_j f(c_{ij})\\

$${#eq-access-singlyconstrained-origin}

\noindent Where $A_i$ and $B_j$ are:
$$
A_i = \frac{1}{\sum_j O_j f(c_{ij})}\\
B_j = \frac{1}{\sum_i P_i f(c_{ij})}\\
$$

In terms of origin-constrained -- ..? .

Regarding destination-constrained, has it's origins are in retail models. e.g., decision model for shops to optimally spatially locate to maximise their potential consumer interaction. Starting with reilly's law of gravitational retail (1931), formalized later in @huffDefiningEstimatingTrading1964 and @lakshmananRetailMarketPotential1965. Spatial availability is explicitly also destination-constrained [@soukhovIntroducingSpatialAvailability2023;@soukhovMultimodalSpatialAvailability2024]. Other literature also demonstrates this constraint implicitly through its analytical formulation; for instance the Intervening Opportunities model [@stoufferInterveningOpportunitiesTheory1940; @stoufferINTERVENINGO PPORTUNITIESCOMPETING1960] "axiomatically" @weibull_axiomatic_1976, competition consideration but as a per population rate [@shen1998], FCA methods in general []

For instance, the most recent example spatial availability is formulated like so:
$$
V_{ij}'' = O_j\frac{F^p_{ij} \cdot F^c_{ij}}{\sum_{i=1}^K F^p_{ij} \cdot F^c_{ij}}
$$
Where the $F^p_{ij}$ Population balancing factor is:
$$
F^p_{ij} = \frac{P_{i\in r}^\alpha}{\sum_{i=1}^K P_{i\in r}^\alpha}
$$
And the $F^c_{ij}$ travel cost balancing factor:
$$
F^c_{ij} = \frac{f(c_{ij})}{\sum_{i=1}^K f(c_{ij})}
$$
It can be reformulated to fit into the singly-constrained (origin case) gravity model as the denominators in both those balancing factors cancel out:
$$
V_{ij}'' = O_j\frac{F^p_{ij} \cdot F^c_{ij}}{\sum_{i=1}^K F^p_{ij} \cdot F^c_{ij}}\\

V_{ij}''^{(1)} =O_j \frac{1}{\sum_j O_j f(c_{ij})} P_i  f(c_{ij})\\
$$

@@

(only when the units of the marginals are the same, therefore of more limited application)
ex. @hornerExploringMetropolitanAccessibility2004 and @allenMeasureCompetitiveAccess2020

@@

>>>>

---
title: "A family of accessibility measures (bringing balance back?)"
format: pdf
bibliography: "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
---

@@

```{r load-packages, include=FALSE, cache=FALSE}
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(flextable)
library(knitr)
library(glue)
library(here)
# library(huxtable) # Easily Create and Style Tables for LaTeX, HTML and Other Formats
# library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax
# library(patchwork) # The Composer of Plots
# library(RColorBrewer) # ColorBrewer Palettes
# library(sf) # Simple Features for R
# library(scales) # Scale Functions for Visualization
# library(shadowtext) # Shadow Text Grob and Layer
# library(skimr) # Compact and Flexible Summaries of Data
# library(spdep) # Spatial Dependence: Weighting Schemes, Statistics
# library(tidyr) # Tidy Messy Data
# library(tmap) # Thematic Maps
# library(ggpmisc) # Miscellaneous Extensions to 'ggplot2'
# library(ggrepel) # Automatically Position Non-Overlapping Text Labels with 'ggplot2'
# library(ggspatial) # Spatial Data Framework for ggplot2
# library(gridExtra) # Miscellaneous Functions for "Grid" Graphics 
# library(here)
```

@@

\noindent Where $B_a$ is the amount of business drawn to $a$ from $T$, $P_a$ and $P_T$ are the populations of the two settlements, and $D_{aT}$ is the distance between them. Quantity $N$ was chosen by Reilly in a somewhat _ad hoc_ fashion as 1, and he used empirical observations of shoppers to choose a value of $n = 2$.

@@

The entropy maximization derivation is demonstrated in detail elsewhere (e.g., @ortuzar_2011_modelling, @wilsonSTATISTICALTHEORYSPATIAL1967). The basis of the approach is to accept that there are macro, meso, and micro states. In the case of trips being the interaction variable, a micro state is an individual trip maker, the meso state is the statistically representative zone representing a collection of individual trip makers, and the macro state is the entire study region. Using these assumptions, a set of $T_{ij}$ that maximizes the number of micro states in each meso state subject to some sensible macro-state constraints (@eq-constraint0-gravitymodel, @eq-constraint1-gravitymodel, and @eq-constraint2-gravitymodel). From this approach, the gravity model can be interpreted as the given the total number of individual trips (micro level) statistically represented at origin and destination zones (meso level), the cost of travelling (micro level, represented at the meso level), and the fixed total cost of travelling in the region (number of trips and cost of travel at the macro level). There is then a most probable distribution of trips between zones (meso-states) that stasis all the macro-state constraints.

@@

And finally, Wilson's balancing factors are interpreted as inverse Hansen-type accessibility, especially in the context of competition [@vickermanAccessibilityAttractionPotential1974; @karstEvaluationAccessibilityImpacts2003;@geurs2006accessibility; @willigers2007accessibility;@el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling]. This interpretation is fraut, upon considering @fotheringhamSpatialCompetitionAgglomeration1985's demonstration, which many of these works do not contend with.

@@

$$
V_{ij} = k W_j^{(2)} f(c_{ij}) = k O_j f(c_{ij})
$${#eq-access}

Where $V_{ij}$ consists of the same variables, however $V_{ij}$ explicitly measures _potential interaction_ as a unit of number of potentially reachable opportunities and the single proportionality constant $k$ ensures the total sum of $V_{ij}$ represents the total potential reachable opportunities in the data. Notably, the mass vectors $W_i^{(1)}$ and $W_j^{(2)}$ are replaced by $P_i$ and $O_j$, representing the weight of the origin (population) and destination (opportunities), the weights common in accessibility literature. As accessibility is the sum of opportunities for a $j$ from $i$, when the mass of the origin ($P_i$) is relevant, it enters @eq-access through the $k$ constant. Accessibility is also most often presented as a locational measure, so the value of $V_i$ is of most interest, e.g., the number of potentially reachable opportunities at a zone:

$$
V_{i} = k \sum_j O_j f(c_{ij})
$${#eq-access-sum}

Like with @eq-phys-gravity-model, we distinguish @eq-access into four cases based on versions of the gravity-model building constraints (@eq-constraint0-access,  @eq-constraint1-access, and @eq-constraint2-access) to reflect macro-level knowledge, i.e., the sum of the number of potential reachable opportunities.

@@

All four cases are differentiated like:
- The unconstrained case: neither @eq-constraint0-access, @eq-constraint1-access, or @eq-constraint2-access hold;
- The total constrained case: only @eq-constraint0-access hold
- The singly-constrained case: either @eq-constraint1-access holds (origin-constrained) <!-- production--> or @eq-constraint2-access holds (destination-constrained)<!-- attraction-->; 
- The doubly-constrained case: both @eq-constraint1-access and @eq-constraint2-access hold simultaneously <!--production-attraction-->

<!-- - HYBRID CASE: @eq-constraint1-access or @eq-constraint2-access partly hold -- hybrid production constrained case <!-- (could be repeated for attraction constrained case... and doubly constrained case)-->

Like gravity model problem, we can solve these cases using spatial weights matrix and travel cost matrix, but with alternative meanings. $P_i$ (our $W_i^{(1)}$) is population at $i$ who seek opportunities. $O_j$ (our $W_j^{(2)}$) are the opportunities being sought. For this simple example, we can say that the opportunities are the population's jobs. In our region, each person has a job - but we only know at an aggregate how many people are at each $i$, the jobs that are located at $j$ and the travel cost $c_{ij}$. We assume the the travel impedance ($f(c_{ij})$) is the associated travel propensity of $f(c_{ij}) = 1/C$.

The following is the known accessibility matrix. The $V_{ij}$, the number of jobs spatially accessible at $i$ from destination $j$. The TOTALs are the $V_i$ and $V_j$, the total number of jobs spatially available at $i$ and the number of jobs spatially available from $j$.
```{r}
known_matrix <- matrix(c(95,23,42, 27,378,45, 38,36,106), ncol = 3, byrow = TRUE)
#unknowns <- matrix(c(rep(NA,9)), ncol=3, byrow=TRUE)
P_i <- known_matrix |> rowSums()
O_j <- known_matrix |> colSums()
V <- ifelse(sum(P_i)==sum(O_j), sum(O_j),NA)
knowns_longer <- cbind(known_matrix,P_i)
knowns_longer <- rbind(knowns_longer,cbind(t(O_j),T))

@@

This accessibility is standard practice. There is no explicit consideration of a proportionality constant ($k$). We can presume it is still set to 1 as @stewartPrinciples1947. No constraints hold, making the interpretation of the units of $V_{ij}$ simply some magnitude of $O_j$ weighted by some $f(c_ij)$, and $V_{i}$ the sum of these values for each $j$. $V_i$ reflects the magnitude of opportunity accessibility but it is not scaled to equal the number opportunities $O$ and/or population $P$.
$$
V_{ij} = O_j f(c_{ij})\\ 
V_{i} = \sum_j O_j f(c_{ij})
$$

Each origin (row) all the opportunities at each destination are available. Likewise, each destination (column) offers all opportunities to each origin: 
```{r}
O_P_unc <- t(matrix(rep(O_j, 3), 3, 3))
O_P_unc_longer <- O_P_unc
rownames(O_P_unc_longer) <- c("1","2","3") 
colnames(O_P_unc_longer) <- c("1","2","3")
O_P_unc_longer
```

The resulting matrix of $V_{ij}$ along with the $V_i$ and $V_j$ totals for this example:
```{r}
V_unc <- O_P_unc / C

P_i_unc <- V_unc |> rowSums()
O_j_unc <- V_unc |> colSums()
V_TOTAL_unc <- ifelse(sum(P_i_unc|>round())==sum(O_j_unc|>round()), sum(V_unc), NA)
V_unc_longer <- cbind(V_unc,P_i_unc)
V_unc_longer <- rbind(V_unc_longer, cbind(t(O_j_unc),V_TOTAL_unc))

rownames(V_unc_longer) <- c("1","2","3", "TOTAL") 
colnames(V_unc_longer) <- c("1","2","3", "TOTAL")

V_unc_longer |> round(digits = 2) #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
```

We'd report origin 1 as having 148 accessibility (i.e., $V_{i=1} = 148$), origin 2 as 248 accessibility, and origin 3 as 172 accessibility. The total accessibility in the region is 568 ($\sum_i V_{i} = 568$), not equal to any meaningful value. This value is not often reported in the accessibility literature for this reason. One could also report the accessibility values that destinations offers ($V_{j}$), namely destination 1 offers 123, destination 2 offer 291, and destination 3 offers 154. However, these values are also not reported in the literature currently... as they aren't constrained to match up to the number of opportunities at each j (i.e., 160, 437 and 193). Assuming $k=1$, $V_i$ and $V_j$ represent magnitudes of accessibility concepts, but in an unconstrained way. The population matrix is no incorporated in the $V_{ij}$ calculation, so $V_i$ cannot be directly compared to population. From a different perspective, $V_j$ values also cannot be compared to known $O_j$ values as the equation unconstrained.

OF NOTE: in the context of spatial interaction modelling, the unconstrained model is not often useful. While the resulting $T_{ij}$ values are in units of 'trips' - the assumption of $k$ equaling 1 is practically meaningless.

Accessibility, or the _potential_ for interaction is different than interaction ($V_{ij}$ vs. $T_{ij}$), namely since the number of opportunities that are accessible at an origin from a destination cannot be known in the same way as a trip from an origin to a destination. However, we would argue that getting closer to knowning this number and planning for it - is what accessibility research should strive to achieve. <!-- yeah? -->

@@

In the total constrained case, $k$ is replaced with $K$ (@eq-access satisfying only constraint @eq-constraint0-access) as follows for $V_{ij}$ and $V_{i}$:
$$
V_{ij}' = K O_j f(c_{ij})\\
V_{i}' = K\sum_j O_j f(c_{ij})
$$

Solved for our example, $K$ is equal to the total accessibility $V$, divided by the sum of the unconstrained accessibility:
$$
K = \frac{V}{\sum_{i}\sum_{j} O_j f(c_{ij})}\\
K = \frac{V}{\frac{O_1}{c_{1,1}}+\frac{O_1}{c_{2,1}}+\frac{O_1}{c_{3,1}}+\frac{O_2}{c_{1,2}} +... \frac{O_3}{c_{3,3}}
}\\
K = \frac{790}{568.4} \\
K = 1.389866
$$

```{r}
K <- V/sum(V_unc)
V_tot <- K * V_unc

P_i_tot <- V_tot |> rowSums()
D_j_tot <- V_tot |> colSums()
V_TOTAL_tot <- ifelse(sum(P_i_tot)==sum(D_j_tot), sum(V_tot), NA)
V_tot_longer <- cbind(V_tot,P_i_tot)
V_tot_longer <- rbind(V_tot_longer, cbind(t(D_j_tot),V_TOTAL_tot))

rownames(V_tot_longer) <- c("1","2","3", "TOTAL") 
colnames(V_tot_longer) <- c("1","2","3", "TOTAL")

V_tot_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
```

While we can see that the 'number of potentially reachable opportunities' `r V` results, the total access at a each $i$ or $j$ is not equal to the known known total access ends (because these constraints are not specified and thus not satisfied by the model). This approach is not yet seen in accessibility literature, but could be an intelligible way to compare accessibility values ($V_i$ or $V_j$) without considering the population matrix.

@@

Similarly, singly-constrained can be either origin-constrained $V_i''^{(1)}$ or destination-constrained $V_i''^{(2)}$. For the origin- and destination- constrained versions, $k$ is replaced with $A_{ij}$ and $B_{ij}$ respectively (@eq-access enforcing constraint @eq-constraint1-access or @eq-constraint2-access, and implicitly @eq-constraint0-access):

$$
V_{ij}''^{(1)} = A_{ij} O_j f(c_{ij})\\
V_{ij}''^{(2)} = B_{ij} O_j f(c_{ij})\\
$${#eq-access-singlyconstrained-origin}

\noindent Where $A_{ij}$ and $B_{ij}$ are similar to the balancing factors of the gravity model for the singly-constrained cases, though with an addition to the numerator to maintain the associated enforcing constraints:
$$
A_{ij} = \frac{O_j}{\sum_j B_{ij} O_j f(c_{ij})}\\
B_{ij} = \frac{P_i}{\sum_i A_{ij} P_i f(c_{ij})}\\
$$
In solving the origin-constrained $V_{ij}^{''(1)}$ case (e.g., @eq-constraint1-access of $\sum_j V_{ij} = P_i$ is enforced), $B_{ij} = 1$. In solving the destination-constrained $V_{ij}^{''(2)}$ case (e.g., @eq-constraint2-access of $\sum_i V_{ij} = O_j$ is enforced), $A_{ij} = 1$.

Solving for the destination-constrained case $V_{ij}^{''(2)}$ may be helpful in an example where the number of people seeking opportunities is known ($P_i$), along with the number of opportunities $O_j$ offered by a destination as in unconstrained and total accessibility. An example could be the number of people at each residence seeking hospital-beds (origin) and hospital-beds (opportunities) at each healthcare location (destination) to calculate the opportunity access at each origin. Another case could be calculating the opportunity access at each origin using the quantity of goods available to purchase (opportunities) at each commercial location (destination).

From our review, related versions of this defined destination-constrained accessibility are formulaically similar to retail accessibility models e.g., decision model for shops to optimally spatially locate to maximize their potential consumer interaction. Starting with reilly's law of gravitational retail [@reilly1929methods], formalized later in @huffDefiningEstimatingTrading1964 and @lakshmananRetailMarketPotential1965. Other literature also demonstrates  similar _competition_ considerations through its analytical formulation; for instance, in the Intervening Opportunities model [@stoufferInterveningOpportunitiesTheory1940; @stoufferINTERVENINGOPPORTUNITIESCOMPETING1960] "axiomatically" @weibull_axiomatic_1976, and competition consideration expressed as a per population rate particularly in healthcare and job opportunity accessibility analysis [@shen1998], and 2FCA methods in general [@soukhovIntroducingSpatialAvailability2023]. Most of these works do not explicitly focus on balancing accessibility units; rather, they center on _competition_ considerations. We suspect this may be why they have been used in case studies examining access to opportunities that are inherently competitive or where the concept of supply and demand resonates. It was only until the recent work of [@soukhovIntroducingSpatialAvailability2023], where the units of accessibility given competition, e.g., the spatial availability, was introduced with the focus on both competition and balanced units. NOTE: See Appendix where I demonstrate how this destination-constrained case is equivalent to spatial availability

Returning to our simple example, the destination-constrained singly-constrained case would yield the following $B_{ij}$ values:
$$
B_{ij} = \frac{P_{i}}{\sum_i P_i f(c_{ij})}\\
B_{1,1} = \frac{P_{i}}{P_1 f(c_{1,1}) + P_2 f(c_{1,2}) + P_3 f(c_{1,3})}\\
...\\
B_{1,1} = 160 *\frac{1}{160/2+450/15+180/5} = 1.09589 \\ 
B_{2,1} = 450 *\frac{1}{160/2+450/15+180/5} =3.082192 \\
B_{3,1} = 180 * \frac{1}{160/2+450/15+180/5} =1.232877 \\
...\\
B_{1,2} = 160 *\frac{1}{160/15+450/2+180/10} =0.630749 \\
B_{2,2} = 450 *\frac{1}{160/15+450/2+180/10} =1.773982 \\
B_{3,2} = 180 * \frac{1}{160/15+450/2+180/10} =0.7095926 \\
...\\
B_{1,3} = 160 *\frac{1}{160/5+450/10+180/2} = 0.9580838\\
B_{2,3} = 450 *\frac{1}{160/5+450/10+180/2} = 2.694611  \\
B_{3,3} = 180 *\frac{1}{160/5+450/10+180/2} = 1.077844 \\
$$
```{r}
B_ij <- matrix(P_i) %*% (1/(t((matrix(P_i))) %*% (1/C))) 
B_ij_longer <- B_ij
rownames(B_ij_longer) <- c("1","2","3") 
colnames(B_ij_longer) <- c("1","2","3") 
B_ij_longer 
```

$V_{ij}$ values:
```{r}
V_dest <- (B_ij) *(V_unc) # V_unc = (1/C)* t(matrix(rep(W_j, 3), 3, 3))

P_i_dest <- V_dest |> rowSums()
O_j_dest <- V_dest |> colSums()

V_TOTAL_dest <- ifelse(sum(P_i_dest)==sum(O_j_dest), sum(V_dest), NA)
V_dest_longer <- cbind(V_dest,P_i_dest)
V_dest_longer <- rbind(V_dest_longer, cbind(t(O_j_dest),V_TOTAL_dest))

rownames(V_dest_longer) <- c("1","2","3", "TOTAL") 
colnames(V_dest_longer) <- c("1","2","3", "TOTAL")

V_dest_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
```

The $O_j$ values are equal to the known opportunities, whereas the $P_i$ values are not. The total accessibility in the region is maintained at 790. Each $V_{ij}$ can be interpreted as the number of opportunities accessible from each destination by anyone in the region all weighted by travel cost of course.

@@

(only when the units of the marginals are the same, therefore of more limited application)
ex. @hornerExploringMetropolitanAccessibility2004 and @allenMeasureCompetitiveAccess2020
<!-- this done'st work.. for now!-->
```{r}
doubly_constrained_gravity_model <- function(
    #CREDIT: I converted the python code from (https://github.com/SadraDaneshvar/Gravity_Model) to R
  P,  # Origin production weights vector (POPULATION)
  O,  # Destination attraction weights vector (OPPORTUNITIES)
  impedance_matrix,  # the resulting travel cost impedance matrix
  error_threshold = 0.001,  # Error threshold for stopping condition
  improvement_threshold = 1e-6  # Improvement threshold for stopping condition
) {

  # Normalize O and D
  sum_P <- sum(P)
  sum_O <- sum(O)
  cat("Checking population, opportunity balancing:\n")
  cat("Population: ", sum_P, "\n")
  cat("Opportunities: ", sum_O, "\n")

  if (sum_P != sum_O) {
    cat("Population and opportunities do not balance, scaling to match.\n")
    if (sum_P < sum_O) {
      P <- P * (sum_O / sum_P)
    } else {
      O <- O * (sum_P / sum_O)
    }
  } else {
    cat("Population, opportunities balancing OK.\n")
  }

  n <- length(P)  # Number of i
  V <- sum(P)  # Total access
  Ai <- rep(1, n)  # Initialize Ai
  Bj <- rep(1, n)  # Initialize Bj

  previous_error <- Inf
  iteration_count <- 0
  stop_reason <- ""

  while (TRUE) {
    iteration_count <- iteration_count + 1

    # Update Ai
    for (i in 1:n) {
      Ai[i] <- 1 / (sum(Bj * O * impedance_matrix[i, ]) + 1e-9)
    }

    # Update Bj
    Bj_new <- rep(1, n)
    for (j in 1:n) {
      Bj_new[j] <- 1 / (sum(Ai * P * impedance_matrix[, j]) + 1e-9)
    }

    # Calculate Vij
    #Vij <- outer(Ai * P, Bj_new * O) * impedance_matrix
    Vij <- Bj_new * Ai * O * impedance_matrix

    # Print Ai and Bj for this iteration
    cat("\nIteration:", iteration_count, "\n")
    cat("Ai:", round(Ai, 3), "\n")
    cat("Bj:", round(Bj_new, 3), "\n")

    # Calculate error
    error <- (sum(abs(P - rowSums(Vij))) + sum(abs(O - colSums(Vij)))) / V

    # Calculate change in error
    error_change <- abs(previous_error - error)

    # Check stopping conditions
    if (error < error_threshold) {
      stop_reason <- "Error threshold met"
      break
    } else if (error_change < improvement_threshold) {
      stop_reason <- "Slow improvement"
      break
    }

    previous_error <- error
    Bj <- Bj_new
  }

  # report stopping condition and final error threshold
  cat("\nStopping Condition: ", stop_reason, "\n")
  cat(sprintf("Final Error: %.3f%%\n", error * 100))

  return(list(Ai = Ai, Bj = Bj, Vij = Vij))
}

dc_results <- doubly_constrained_gravity_model(P=P_i, O=O_j, impedance_matrix = 1/C)
dc_results$Vij
```

The $A_i$ and $B_j$ matrices:
```{r}
A_i_doubly <- dc_results$Ai
B_j_doubly <- dc_results$Bj

A_i_doubly |> matrix()
B_j_doubly |> matrix() |> t()
```

The trip matrix (notice, all summed up opportunities and populations are equivalent to our knowns)
```{r}
V_doubly <- dc_results$Vij

P_i_doubly <- V_doubly |> rowSums()
O_j_doubly <- V_doubly |> colSums()
V_TOTAL_dest <- ifelse(sum(P_i_doubly)==sum(O_j_doubly), sum(V_doubly), NA)
V_doubly_longer <- cbind(V_doubly,P_i_dest)
V_doubly_longer <- rbind(V_doubly_longer, cbind(t(O_j_doubly),V_TOTAL_dest))

rownames(V_doubly_longer) <- c("1","2","3", "TOTAL") 
colnames(V_doubly_longer) <- c("1","2","3", "TOTAL")

V_doubly_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
```

@@

# APPENDIX (some musings...)

## Singly-constrained case is spatial availability

The destination-constrained case is also spatial availability, see below! As follows then, this case is also related to 2SFCA - if the value is divided by per capita (OR $P_i$ from the numerator in $B_{ij}$ is removed).

Spatial availability is formulated like so:
$$
V_{ij}''^{(2)}  = O_j\frac{F^p_{ij} \cdot F^c_{ij}}{\sum_{i=1}^K F^p_{ij} \cdot F^c_{ij}}
$$
It turns out that the denominators of the Population balancing factor $F^p_{ij} = \frac{P_{i\in r}^\alpha}{\sum_{i=1}^K P_{i\in r}^\alpha}$ and Travel cost balancing factor $F^c_{ij} = F^c_{ij} = \frac{f(c_{ij})}{\sum_{i=1}^K f(c_{ij})}$ cancel out, so spatial availability can be expressed as:
$$
V_{ij}''^{(2)} =O_j \frac{ P_i  f(c_{ij})}{\sum_i P_i f(c_{ij})}\\
$$

Reformatted in the general family of accessibility measure form, it looks like:
$$
V_{ij}''^{(2)} =k O_j f(c_{ij})\\
V_{ij}''^{(2)} =\frac{ P_i}{\sum_i P_i f(c_{ij})} O_j f(c_{ij})\\
V_{ij}''^{(2)} =  O_j\frac{F^p_{ij} \cdot F^c_{ij}}{\sum_{i=1}^K F^p_{ij} \cdot F^c_{ij}} = O_j \frac{ P_i  f(c_{ij})}{\sum_i P_i f(c_{ij})}\\
$$
## Checking singly-constrained case against spatial availability
```{r}
sp_avail <- function(x, o_id, d_id, pop, opp, r, f, alpha = 1){

  o_id <- rlang::enquo(o_id)
  d_id <- rlang::enquo(d_id)
  pop <- rlang::enquo(pop)
  opp <- rlang::enquo(opp)
  r <- rlang::enquo(r)
  f <- rlang::enquo(f)

  sum_pop <- x %>%
    dplyr::distinct(!!o_id,
                    .keep_all = TRUE) %>%
    dplyr::mutate(sum_pop = !!r*(!!pop)^alpha) %>%
    dplyr::pull(sum_pop) %>%
    sum()

  f_p <- dplyr::pull(x, !!r) * dplyr::pull(x, !!pop)^alpha / sum_pop

  sum_impedance <- x %>%
    dplyr::group_by(!!d_id) %>%
    dplyr::summarize(sum_impedance = sum(!!f))

  x <- x %>%
    dplyr::left_join(sum_impedance,
                     by = rlang::as_name(d_id))

  f_c <- dplyr::pull(x, !!f) / x$sum_impedance

  x$f_c <- f_c
  x$f_p <- f_p

  sum_pa <- x %>%
    dplyr::group_by(!!d_id) %>%
    dplyr::summarize(sum_pa= sum(f_p * f_c))

  x <- x %>%
    dplyr::left_join(sum_pa,
                     by = rlang::as_name(d_id))
  f_t <- (f_p * f_c) / dplyr::pull(x, sum_pa)

  dplyr::pull(x, !!opp) * f_t
}
```

```{r}

df <- data.frame(i = c(1,1,1,2,2,2,3,3,3), 
                 j= c(1,2,3,1,2,3,1,2,3),
                 f = c(1/C[1,], 1/C[2,], 1/C[3,]),
                 pop = c(P_i[1], P_i[1], P_i[1], P_i[2], P_i[2], P_i[2], P_i[3], P_i[3], P_i[3]),
                 opp = c(O_j[1],O_j[2],O_j[3], O_j[1],O_j[2],O_j[3], O_j[1],O_j[2],O_j[3]),
                 catch = c(1,1,1,1,1,1,1,1,1))

calced <- sp_avail(df,
                         o_id = i,
                         d_id = j,
                         pop = pop,
                         opp = opp,
                         r = catch,
                         f = f)
df$V_ij <- calced

df |> group_by(j) |> summarise(V_j = sum(V_ij))
```

## Destination- (or ATTRACTION) constrained is the only singly-constrained case... population- (or PRODUCTION) constrained doesn't make sense!

Then the origin-constrained (singly-constrained) case $V_{ij}^{(1)}$, no $P_i$ values are needed only $O_j$ values. Using this method, V_i values equal P_i values...? while O_j values are unconstrained. This case may be helpful in an example where the number of opportunities available to an origin is known and fixed. This could be a normative minimum threshold, such as 100 opportunities at each origin.
